% !BIB program = biber
\documentclass[stu]{apa7}

\raggedbottom

\usepackage[american]{babel}
\usepackage{lipsum}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage[style=apa,backend=biber]{biblatex}
\usepackage{amssymb}
\addbibresource{ppb.bib}
% \bibliographystyle{unsrt} % Or use a specific apa-numeric style
% \bibliography{ppb.bib}

%% --- Proposal Metadata ---
\title{Physics Informed Hyperspectral Image Reconstruction from RGB}
\shorttitle{Project Proposal} % Required for APA headers

\authorsnames{Aditya Bhargava}
\authorsaffiliations{University of Georgia}
\course{CSCI 8820: Computer Vision and Pattern Recognition}
\professor{Dr. Suchendra Bhandarkar}
\duedate{March 18, 2026}

%% --- Abstract (Proposal Executive Summary) ---
\abstract{Hyperspectral image (HSI) reconstruction from single-shot RGB measurements has predominantly relied on purely data-driven deep learning, often treating the complex physical imaging process as a black-box (three to many) mapping problem. Since this is an ill-posed problem it leads to difficulties, particularly in distinguishing similar materials due to the metamer problem—where distinct physical spectral signatures project to identical RGB values. To address this fundamental limitation, this proposal outlines a Physics-Informed architecture. Moving from direct RGB-to-HSI mapping, the proposed framework structurally embeds the camera's Spectral Response Function (SRF) to compute a mathematically guaranteed physical baseline. A deep neural network is then utilized to model the spectral discrepancy, explicitly tasked with learning the invisible metameric black component residing in the SRF null-space. To ensure physical validity and prevent shape overfitting, the system incorporates metameric data augmentation, a differentiable forward-model constraint layer, and an adaptively weighted multi-task loss function. The expected deliverable is a robust reconstruction pipeline capable of maintaining strict spectral reconstruction under metameric conditions, effectively bridging the gap between computational deep learning and the physical principles of hyperspectral data acquisition.}

\keywords{Computer Vision, HSI, RGB, Physics Informed}

\begin{document}
\maketitle



\section{Introduction and Problem Statement}

% Describe the "Why". What is the gap in current computer vision research or applications?
Hyperspectral imaging (HSI) is an advanced optical sensing technique that assembles spectroscopy and digital photography into a single system  \parencite{hong2026hyperspectralimaging}. The imaging collects and processes scene information by dividing the whole spectrum into tens of hundreds of bands, typically covering wavelengths from 380 nm to 2500 nm \parencite{Rybicki:847173}. The technique has been widely adopted in many fields, e.g., medical diagnosis, healthcare, remote sensing and identifying foreign objects in food products \parencite{ghimpeteanu2025hyperspectralimagingidentifyingforeign}. The hyperspectral images possess a three-dimensional structure, comprising two dimensions of spatial information and one dimension of spectral information. However, conventional spectrometers often operate in 1D line or 2D plane scanning, which is time-consuming and not suitable for dynamic scenes. Also, the devices for acquiring the HSIs are typically more complicated and expensive than common cameras. 
% In order to obtain the spectral information more effectively by extracting the required spectral information from the RGB images. 
The need to reconstruct the spectral properties of scene reflectance (and illumination) from a three-channel device or standard color space was recognized as early as 1980s \parencite{Hall1988IlluminationAC,Jaaskelainen:90, Maloney:86, Maloney:86, Parkkinen:89}. But this is an extremely ill-posed problem since a large amount of information is lost during the process of spectral integration when RGB sensors capture the light.\\ Spectral reconstruction from RGB images often suffer from a \textit{metameric} dilemma, where distinct spectral distribution map to nearly identical RGB values, making them indistinguishable. Recent approaches involve deep learning methods to solve this by learning the mapping from RGB to HSI. While these models achieve reasonable quantitative metrics, they operate as black boxes that ignore the underlying mechanism behind the imaging process. To overcome these issues, the goal is to incorporate physics based formulations into the architecture itself. By explicitly embedding the camera's Spectral Response Function (SRF) as a mathematical forward-model constraint, the network is forced to output hyperspectral reconstructions that strictly adhere to the physical laws of image formation, achieving both higher accuracy and reliable material distinction.

\section{Literature Review}
A RGB image is usually generated in three wide spectral channels with Bayer Filters, which is essentially a color filter array (CFA) placed over digital image sensor to capture color information. While, an HSI is obtained by the interaction between the spectral reflectance and illumination. The relation between the RGB and HSI images can be represented by

\begin{equation}
    I_c(x, y) = \int_{w_1}^{w_2} R(x, y, w) L(w) S_c(w) \, dw
\end{equation}
where $I_c$ represents each channel of RGB image, $S_c(w)$ denotes the spectral response function and $R(x,y,z)$ is the spectral reflectance of the objects and $L(w)$ is the illumination spectrum. HSI is typically defined as
\begin{equation}
    I_c(x, y) = \int_{w_1}^{w_2} H(x, y, w) S_c(w) \, dw
\end{equation}
where, $H(x,y,z)$ represents the Hyperspectral image. The challenge in spectral reconstruction is to invert the above forward image formation model in optimization framework to find the $H(x, y, z)$. A wide variety of methods have been developed to achieve spectral reconstruction.

\subsection{Prior-Based Methods}
These methods rely on mathematical assumptions around which the algorithms are designed. Broadly there are three categories of Prior-Based methods namely dictionary learning, manifold learning and gaussian process. \textcite{Arad2016SparseRO} introduced a method to reconstruct HSI directly from RGB leveraging hyperspectral priors in order to create a sparse dictionary of hyperspectral signatures and their corresponding RGB projections. The method works by collecting the hyperspectral prior, preferably from a set of domain specific scenes. This prior is then reduced computationally to an overcomplete dictionary $D_h$ of hyperspectral signatures. The dictionary is projected to the sensor space via the receptor spectral absorbance functions $R$, establishing a corresponding RGB dictionary, $D_{rgb}$. 
This correspondence between each RGB vector and its hyperspectral signature originator $h_i$ is maintained for the later mapping from RGB to hyperspectral signatures. For a given image, a sparse weight vector $w$ is calculated, that reconstructs a given RGB pixel query from the RGB dictionary. Because both dictionaries share the exact same sparse coefficients, applying these weights back to the hyperspectral dictionary yields the estimated spectrum signature $h_q$. 

While, manifold learning approach is aimed at finding a low-dimensional manifold that can uniquely represent the high dimensional data. \textcite{8237766} proposed a manifold approach but simplifying the three to many mapping into a well conditioned three to three problem using a isometric mapping algorithm which is used to reduce dimensionality of the basis spectra to a three-dimensional subspace. After finding a low dimensional embedding of natural scene spectra, a non linear mapping is learned between the RGB vectors and their corresponding 3D embedded natural scene spectra using a radial basis function with 10 hidden neurons and leveraging Levenberg-Marquardt training algorithm \parencite{Levenberg1944AMF,Marquardt1963AnAF}. The reconstruction of high dimensional spectrum from its 3D embedding employs a dictionary learning-based technique, essentially finding a common coding between the 3D embedding and high dimensional space of the spectra.

\textcite{akhtar2018hyperspectralrecoveryrgbimages} proposed a technique to estimate the HSI from RGB by modeling natural spectra with Gaussian processes. By exploiting the kernels of gaussian process to model the relative smoothness of naturally occurring spectra and accounting for the non-negativity of reflectance values and the self-similarity of hyperspectral image patches for accurate spectral recovery.
% this formulation inherently guarantees that the reconstructed hyperspectral signal is physically consistent with the original RGB measurement (i.e., $c_q = R \cdot h_q$), though the overall reconstruction quality remains strictly bounded by the representational capacity of the initial hyperspectral prior.

\subsection{Deep-Learning Methods}
\cite{Xiong2017HSCNNCH} introduced a deep learning approach to recover the hyperspectral image. The method first applies a upsampling matrix to interpolate the 3-channel input into a high-dimensional spectral volume. A convolutional network, inspired by the residual learning architectures like VSDR, is then applied to the the volume. The network learns a learns and end-end to mapping from a large number of upsampled/ground truth hyperspectral image pairs.

Building upon the original HSCNN, the advanced HSCNN+ \parencite{Shi2018HSCNNAC} models eliminate the need for hand-crafted upsampling and significantly improve feature extraction.
Instead of using a fixed interpolation matrix, HSCNN+ utilizes a pointwise convolutional layer to directly expand the 3-channel RGB input into the targeted spectral dimensions and replacing the plain convolution layers in HSCNN by modern residual blocks to boost its performance allowing the the network to learn the optimal spectral mapping.

\textcite{cai2022mstmultistagespectralwisetransformer} introduced Vision Transformer based architecture shifting from CNN-based methods. Excelling at capturing long-range dependencies in hyperspectral data, MST$++$ employs a novel spectral-wise multi head self-attention (S-MSA) that is based on the hyperspectral characteristics. Which allows the model to capture global spatial context and complex spectral correlations simultaneously, which traditional CNNs struggle to achieve due to their localized receptive fields
\section{Proposed Methodology}
% How are you going to build it?

\subsection{Proposed Algorithm/Model Architecture}

To bridge the gap between deep learning methods and the physical principles of the imaging process/data acquisition, I propose a Physics-Informed framework. Moving beyond standard mapping by embedding physical priors and utilizing metameric data augmentation. It consists of a physical baseline estimation, a deep feature extraction backbone for discrepancy modeling, and a differentiable forward-model constraint layer.

\subsubsection{Training Data Curation: Metameric Augmentation}
Hyperspectral datasets lack sufficient metameric diversity and generally have fewer training instances, causing networks to easily overfit to brightness and shape. To mitigate this, I utilize metameric augmentation \parencite{romero2025uncertaintyquantificationhsireconstruction} step during training. Following the metameric black theory, the ground truth spectrum $H_{gt}$ can be decomposed into a fundamental metamer $S^*$ and an invisible metameric black component $B$. This metameric black component exists within the null-space of the camera's Spectral Response Function (SRF) projection matrix $\Phi$, mathematically satisfying the condition $B \Phi = 0$. Because of this property, adding any amount of $B$ does not alter the resulting RGB projection. A physically valid augmented ground truth $H'_{gt}$ is generated by scaling the metameric black by a random coefficient $\alpha$:
$$H'_{gt} = S^* + \alpha B$$
This step trains the network to explicitly recognize the underdetermined nature of the metamer problem.

\subsubsection{Discrepancy Modeling Network}
Instead of simply mapping from RGB to a full $C$-channel hyperspectral volume from scratch, the pipeline adapts the discrepancy modeling framework to resolve the model-measurement mismatch in the spectral domain. First, an approximate physical baseline $H_{baseline} \in \mathbb{R}^{H \times W \times C}$ is calculated using the pseudo-inverse of the camera's Spectral Response Function (SRF), denoted as $\Phi^{\dagger}$:
$$H_{baseline} = I_{rgb} \Phi^{\dagger}$$
Mathematically, this least-squares minimum-norm solution is equivalent to the fundamental metamer $S^*$ defined in the previous section. Because this linear approximation cannot capture complex physical textures, the deep neural network (acting as the discrepancy mapping function $f_\theta$) is utilized to identify the missing physics. Consequently, the network is tasked exclusively with learning the missing metameric black component (the null-space residual). The network takes $I_{rgb}$ and learns a model for the systematic state-space residual, predicting only the high-frequency residual details $\Delta H \in \mathbb{R}^{H \times W \times C}$. I plan to evaluate reliable backbones such as U-Net or ResNet (as well as lightweight Vision Transformers) for this stage. The final predicted hyperspectral volume is the sum of the physical baseline and the learned discrepancy:
$$\hat{H} = H_{baseline} + \Delta H$$

\subsubsection{Differentiable Forward Model Layer}
A parameter-free, differentiable layer mathematically simulates the camera's sensor integration. By discretizing the forward model, I project the predicted hyperspectral image $\hat{H}$ back into the RGB domain to generate a simulated RGB image, $\hat{I}_{rgb}$, using the known SRF projection matrix $\Phi \in \mathbb{R}^{C \times 3}$:
$$\hat{I}_{rgb} = \hat{H} \Phi$$
This layer acts as a strict filter. The network is explicitly forced to learn how its spectral output will be integrated by the specific camera sensor.

\subsubsection{Loss Function with Adaptive Weighting}
To train the network, I treat the joint optimization as a dynamic multi-task learning problem, evaluating both the augmented data prior and physical inconsistencies. The total loss function $\mathcal{L}_{total}$ is defined as:
$$\mathcal{L}_{total} = \mathcal{L}_{recon}(\hat{H}, H'_{gt}) + \lambda \mathcal{L}_{physics}(\hat{I}_{rgb}, I_{rgb})$$
where $\lambda$ is an adaptively weighted hyperparameter. This is to prevent one task loss from dominating the other entirely.

\begin{itemize}
    \item \textbf{Data Prior Loss ($\mathcal{L}_{recon}$):} I will utilize the $L1$ Loss (Mean Absolute Error), as standard MSE struggles with metameric gradients, to ensure the predicted spectrum closely matches the augmented ground truth $H'_{gt}$.
    \item \textbf{Physics Constraint Loss ($\mathcal{L}_{physics}$):} Computed between the input RGB image $I_{rgb}$ and the simulated RGB image $\hat{I}_{rgb}$. Minimizing this mathematically restricts the network's prediction to a valid metamer under the specific SRF constraint.
    \item \textbf{Adaptive Weighting:} To prevent dominant terms from overshadowing others, the optimizer will dynamically adjust the weight $\lambda$ based on the gradient statistics of the neural tangent kernel (NTK) or gradient path non-uniformity. This balances the convergence rates of the different losses without manual hyperparameter tuning.
\end{itemize}

\subsection{Evaluation Metrics}
To comprehensively evaluate the reconstruction accuracy and physical validity, the following metrics will be utilized: Mean Relative Absolute Error (MRAE), Root Mean Square Error (RMSE), Peak Signal-to-Noise Ratio (PSNR), and Spectral Angle Mapper (SAM) to specifically evaluate spectral shape fidelity. Computational performance will be evaluated based on model parameter count and inference time.

\subsection{Data Acquisition and Preprocessing}

To train and evaluate the proposed physics-informed framework, hyperspectral datasets are required to serve as the ground truth $H_{gt}$. I will utilize widely adopted public benchmarks, such as the Interdisciplinary Computational Vision Laboratory (ICVL) dataset and the CAVE database. These datasets contain a diverse array of spectral profiles capturing both natural and artificial materials, which is critical for evaluating metameric robustness of the network.

\subsubsection{Input Pair Generation via Known SRF}
To train the model, perfectly aligned pairs of RGB inputs ($I_{rgb}$) and hyperspectral ground truths ($H_{gt}$) are required. Because physical RGB cameras introduce a non-linear Image Signal Processor (ISP) distortions (e.g., gamma correction, white balance) that violate strict linear spectral integration could be induced, the RGB inputs must be mathematically derived. 

Using the known, camera Spectral Response Function (SRF), denoted as $\Phi \in \mathbb{R}^{31 \times 3}$, the high-dimensional spectral cubes ($C=31$) are projected into the RGB domain:
$$I_{rgb} = H_{gt} \Phi$$
By generating the inputs via this method using standard SRFs (e.g., CIE 1931 or a calibrated DSLR response curve) while taking the dataset into consideration, it guarantees that the physical baseline calculation ($I_{rgb} \Phi^{\dagger}$) and the forward physics loss remains free from any unknown nonlinearities.

\subsubsection{Preprocessing Pipeline}
Prior to training the neural network, the raw hyperspectral data will undergo a preprocessing pipeline to ensure physical consistency, stability, and computational efficiency:

\begin{itemize}
    \item \textbf{Spatial Patching:} The high-resolution spatial dimensions of the hyperspectral cubes will be divided into overlapping spatial patches (e.g., $128 \times 128$ or $256 \times 256$ pixels). This operation significantly augments the total number of training samples and maintains a manageable memory footprint during the calculation of the pseudo-inverse baseline and backpropagation.
    \item \textbf{Radiometric Normalization:} To ensure gradient stability across the multi-task loss function \parencite{chen2018gradnormgradientnormalizationadaptive}, both the hyperspectral volumes $H$ and the simulated RGB inputs $I_{rgb}$ will be strictly normalized to a continuous $[0, 1]$ floating-point range. 
    \item \textbf{Data Partitioning:} The patched dataset will be partitioned into strictly non-overlapping training, validation, and testing sets. Care will be taken to split the data at the scene level, rather than the patch level, to guarantee that no spatial or spectral data leakage occurs between the training and testing phases.
\end{itemize}

\section{Project Plan and Timeline}
As this is an individual term project, the execution is divided into four phases. The timeline prioritizes the sequential implementation of the physics-informed architecture, starting with data pipeline construction, followed by model training and metameric evaluation. The milestones are detailed in Table~\ref{tab:Timeline}.

\begin{table}[h]
  \centering
  \caption{Project Milestone Timeline}
  \label{tab:Timeline}
  \begin{tabular}{@{}lp{8cm}l@{}} \toprule
  Phase & Task Description & Completion Date \\ \midrule
  1     & \textbf{Data Pipeline \& Augmentation Setup:} Download datasets (ICVL/CAVE), implement the SRF RGB projection ($I_{rgb} = H_{gt} \Phi$), and code the PyTorch 'DataLoader' with metameric augmentation ($H'_{gt} = S^* + \alpha B$). & March 24, 2026 \\ \addlinespace
  2     & \textbf{Architecture Setup \& Initial Training:} Implement the physical baseline calculation ($H_{baseline} = I_{rgb} \Phi^{\dagger}$), build the discrepancy network backbone (ResNet/U-Net), and integrate the differentiable forward model layer. Run initial training loops. & April 7, 2026 \\ \addlinespace
  3     & \textbf{Loss Optimization \& Evaluation:} Integrate the adaptive weighting algorithm for the dual-loss function. Conduct final testing and calculate quantitative metrics (MRAE, RMSE, SAM) to evaluate metameric robustness. & April 17, 2026 \\ \addlinespace
  4     & \textbf{Final Report Formulation:} Compile quantitative results, visual spectral curve comparisons, finalize the write-up, and format the final technical report for submission. & April 27, 2026 \\ \bottomrule
  \end{tabular}
\end{table}

\section{Expected Outcomes and Impact}

The proposed Physics-Informed framework aims to improve the reliability of hyperspectral image (HSI) reconstruction. By explicitly separating the known optical physics from the learned state-space residuals, this project targets the following practical outcomes:

\subsection{Technical and Performance Outcomes}
\begin{itemize}
    \item \textbf{Improved Spectral Fidelity and Physical Consistency:} Standard data-driven models often struggle with metameric ambiguity, leading to spectral hallucinations. By structurally isolating the metameric black component ($B$) into the null-space residual learned by the network, the proposed model is expected to reduce Spectral Angle Mapper (SAM) errors and produce outputs that better adhere to the forward measurement constraint ($I_{rgb} = \hat{H} \Phi$).
    
    \item \textbf{Training Efficiency and Stability:} Since the deep neural network only needs to approximate the high-frequency ($\Delta H$), I anticipate that combining this with NTK-based adaptive loss weighting will result in a more stable training and potentially reduce the need of an over-parameterized model.
\end{itemize}

\printbibliography

\appendix

% \section{Resource Requirements}
% \label{app:resources}
% % List hardware (GPUs), datasets, or software libraries needed.
% \begin{figure}[h]
%     \caption{Proposed System Architecture Flowchart.}
%     \figurenote{This diagram represents the data pipeline from input sensor to classification.}
%     \label{fig:Arch}
% \end{figure}

\end{document}